---
title: "Rcpp vs pure R performance"
author: "Phillip Labuschagne"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(yasss)
```

To improve performance, various parts of the simulations were implemented in
C++ and accessed via Rcpp. This vignette explains some of the weirder things
that had to be done to get this working and presents a detailed performance
comparison.

## Passing Rcpp functions into Rcpp functions.

Due to the way the mutators and fitness_evaluators are specified in the sim_pop
call, we need to check that you can actually pass functions to Rcpp and that
the performance hits are not too large.

To test this, write a silly function in Rcpp and pass it to another silly
function in Rcpp.

R performs poorly when you grow data structures or when you update a data.frame
repeatedly. So write a C++ function that produces a fibonacci-like sequence and then
write another function that repeatedly call s this fibonacci-like sequence generator
and populates a data.frame with it. Instead of adding the two previous numbers
to each other add the imediately preceding number to the 4th power root of the
number preceding it. Making this modification keeps the number sizes under
control. This will be a good indicator of the performance characteristics for
the kinds of calculations in yass.

### Pure R

```{r}
fibo <- function(s1, s2, n){
  result <- c(s1, s2)
  for (i in 3:n){
    result <- c(result, result[i-2]^0.25 + result[i-1])
  }
  result
}

fibo_df <- function(n, n_columns = 20){
  x <- data.frame()
  for (i in 1:n_columns){
    x[,paste("V", i, sep = '')] <- numeric(0)
  }
  s1 <- 1
  s2 <- 1
  for (i in 1:n){
    y <- fibo(s1, s2, n_columns+2)
    y <- matrix(y[3:(n_columns+2)], ncol = n_columns)
    y <- as.data.frame(y)
    x <- rbind(x, y)
  }
  x
}

print('Time required to compute 50000 elements of the sequence')
start_time <- proc.time()
x <- fibo(1, 1, 5000)
print(proc.time() - start_time)

print('Time required to compute 400 x 400 matrix')
start_time <- proc.time()
y <- fibo_df(40, 40)
print(proc.time() - start_time)
```

### Pure C++

```{r}
print('Time required to compute 50000 elements of the sequence')
start_time <- proc.time()
x <- cpp_fibo_exported(1, 1, 5000)
print(proc.time() - start_time)

#print('Time required to compute 10000 x 10000 matrix')
#start_time <- proc.time()
#y <- yasss:::cpp_fibo_df(1000, 1000)
#print(proc.time() - start_time)
```

### Passing Rcpp into Rcpp

```{r}
#print('Time required to compute 10000 x 10000 matrix')
#start_time <- proc.time()
#x <- yasss:::cpp_fibo_df_pass(1000, 1000, cpp_fibo_exported)
#print(proc.time() - start_time)
```

*IT WORKS !!!!!*

So this means that it is a valid strategy to write a C++ function for the mutator / fitness_evaluator and then pass that function, via R, into another C++ function (sim_next_gen / sim_pop).

## Mutators



## Template boilerplate, keep for now for easy ref.

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
